---
title: "Are the digital categories redundant with the Funktionenplan?"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Are the digital categories redundant with the Funktionenplan?}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 5
)
```

## Background

Every budget line in `digitalhaushalt` carries two independent
classifications:

- **`funktion`** — the official 3-digit code from the federal
  *Funktionenplan*, describing the *political purpose* of the expenditure
  (e.g. defence, health, education). Present for all lines.
- **`kategorie`** — the research project's *digital theme category* (1–9),
  describing the *type* of digitalisation. Present only for confirmed
  digital lines (`digi_klasse` 1–8).

If the two largely overlapped, digital spending could simply be identified
from the Funktionenplan — making the carefully collected `kategorie`
redundant. This vignette tests that, at both levels of the Funktionenplan
hierarchy.

```{r load}
library(DHaushaltR)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Concentration at Hauptfunktion level

For each digital category, we compute what share of its budget lines fall
into the single most common Hauptfunktion (first digit of `funktion`,
stored in `funktion_lvl1`). A share near 100 % means the Funktionenplan
already captures that category; broad spread means the category adds
independent information.

```{r concentration}
digi <- digitalhaushalt |>
  filter(!is.na(kategorie)) |>
  mutate(kat_name = kat_labels[as.character(kategorie)])

concentration <- digi |>
  group_by(kategorie, kat_name) |>
  summarise(
    n         = n(),
    hf_n      = n_distinct(funktion_lvl1),
    hf_share  = max(table(funktion_lvl1)) / n(),
    of_n      = n_distinct(funktion_lvl2),
    of_share  = max(table(funktion_lvl2)) / n(),
    .groups   = "drop"
  ) |>
  arrange(desc(n))

concentration |>
  transmute(
    Category          = kat_name,
    Lines             = n,
    `HF classes (n)`  = hf_n,
    `Top-HF share`    = scales::percent(hf_share, accuracy = 1),
    `OF classes (n)`  = of_n,
    `Top-OF share`    = scales::percent(of_share, accuracy = 1)
  ) |>
  knitr::kable(caption = paste(
    "Concentration of digital categories in the Funktionenplan.",
    "Top share = fraction of lines in the single most common class."
  ))
```

## Does the finer Oberfunktion level help?

The dumbbell plot below shows the top-class share for each category at
both Funktionenplan levels. If going finer rescued the Funktionenplan
(i.e. made categories look more redundant), the Oberfunktion points would
move *right*. They don't.

```{r dumbbell, fig.alt = "Dumbbell chart comparing concentration at Hauptfunktion vs Oberfunktion level for each digital category."}
dmb <- concentration |>
  transmute(
    kat_name,
    `Hauptfunktion (broad)` = hf_share,
    `Oberfunktion (fine)`   = of_share
  ) |>
  pivot_longer(
    c(`Hauptfunktion (broad)`, `Oberfunktion (fine)`),
    names_to  = "Level",
    values_to = "share"
  ) |>
  mutate(kat_name = factor(kat_name, levels = rev(concentration$kat_name)))

ggplot(dmb, aes(share, kat_name)) +
  geom_line(aes(group = kat_name), colour = "grey70", linewidth = 1) +
  geom_point(aes(colour = Level), size = 3) +
  geom_vline(xintercept = 0.5, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1)
  ) +
  scale_colour_manual(
    values = c("Hauptfunktion (broad)" = "#2166ac",
               "Oberfunktion (fine)"   = "#d6604d")
  ) +
  labs(
    title   = "Does a finer Funktionenplan level reduce the spread?",
    subtitle = "Share of lines in the top-ranked Funktionenplan class, per category",
    x       = "Share of lines in most common class",
    y       = NULL,
    caption = "Source: DHaushaltR — Digitalhaushalt Open Data (CC BY-SA 4.0)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

## Key findings

- **Genuinely redundant** (both levels concentrated): `Bundeswehr`
  (98 % in OF 03 *Verteidigung*), `Gesundheitswesen` (72 % in OF 31),
  `Forschung & Innovation` (63 % in OF 16). The Funktionenplan alone
  suffices for these categories.

- **Apparent redundancy, artefact of coarse resolution**: `Digitale
  Kompetenzen` appears concentrated at Hauptfunktion level (93 % in HF 1,
  *Bildung*), but scatters to a top-OF share of only 28 % — the broad
  "education" umbrella hides large internal variation.

- **Genuinely cross-cutting** (dispersed at both levels): `Digitale
  Verwaltung` (2 236 lines, top-HF 56 %, top-OF only 36 % across
  29 Oberfunktionen), `Digitale Wirtschaft`, `Kultur & Zivilgesellschaft`.
  The Funktionenplan is effectively blind to these categories.

**Conclusion**: the Funktionenplan cannot substitute for the `kategorie`
column. Using both simultaneously as independent predictors is appropriate
for cross-cutting categories; for the concentrated ones (Bundeswehr,
Gesundheit, Forschung) treat them as collinear.
